\documentclass[12pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[a4paper,total={150mm,240mm}]{geometry}
\usepackage[american]{babel}

\usepackage{float}
\usepackage{babel}
\usepackage{amsmath}
\usepackage{tikz}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{wrapfig}
\usepackage{todonotes}
\usepackage{hyperref}

\usepackage{listings}
\definecolor{listingbg}{gray}{0.95}
\lstset{
  language=C++,
  basicstyle=\ttfamily\small,
  frame=single,
  backgroundcolor=\color{listingbg},
  breaklines=true,
  postbreak=\raisebox{0ex}[0ex][0ex]{\ensuremath{\color{red}\hookrightarrow\space}}
  }


\newcommand{\vx}{\vec x}
\newcommand{\grad}{\vec \nabla}
\newcommand{\wind}{\vec \beta}
\newcommand{\Laplace}{\Delta}
\newcommand{\mycomment}[1]{}


% Exercise stylesheet
\usepackage{exercise}

\title{\textbf{Exercises for Tutorial04}}
\subtitle{Finite Elements for the Wave  Equation}
\exerciselabel{Exercise}

\begin{document}

\exerciseheader

In this exercise you will:
\begin{itemize}
\item Work on the problem presented in \lstinline{tutorial04}.
\item Try various time integrators, in particular the Crank-Nicolson
  method.
\item Explore polynomial degrees greater than $2$ (remember to change the
  blocking to \lstinline{none}!).
\item Implement your own local operator that switches to an elliptic
  projection method (see the tutorial notes and Eriksson et
  al. \cite{Eriksson})
\item Check the energy conservation property in the numerical scheme.
\end{itemize}

\begin{Exercise}{Getting to Know the Code}
  \lstset{language=bash} The code of \textbf{exercise04} solves the
  wave equation formulated as a first order in time system. As already
  explained in \lstinline{tutorial04}, we can write the wave equation
  as a system of two equations by substituting $u_0=u$ and introducing
  $u_1=\partial_t u_0 =\partial_t u$:
  \begin{subequations}
    \label{eq:SystemForm1}
    \begin{align}
      \partial_t u_1 - c^2\Delta u_0 &=0 &&\text{in $\Omega\times\Sigma$}, \label{eq:2a}\\
      \partial_t u_0 - u_1 &=0 &&\text{in $\Omega\times\Sigma$}, \label{eq:2b}\\
      u_0 &= 0 &&\text{on $\partial\Omega$},\\
      u_1 &= 0 &&\text{on $\partial\Omega$},\\
      u_0 &= q &&\text{at $t=0$},\\
      u_1 &= w &&\text{at $t=0$}.
    \end{align}
  \end{subequations}
  Since $u_0=u=0$ on the boundary we also have
  $\partial_t u = u_1 = 0$ on the boundary.  Alternatively, one may
  also simply omit the boundary condition on $u_1$.


  The code for this exercise can be recompiled individually \textbf{in
    your build directory} by typing make:
  \begin{lstlisting}
[user@localhost]$ cd release-build/dune-pdelab-tutorials/tutorial04/exercise/task
[user@localhost]$ make clean && make
  \end{lstlisting}

  Note that in contrast to earlier exercises, three different
  executables are built from the same source files, one each for 1D,
  2D and 3D, in order to reduce compile times and compiler memory
  usage.

  The structure of the code is very similar to the previous tutorials
  and consists of the following files:
  \begin{itemize}
  \item \lstinline!exercise04.ini! -- holds parameters read by various
    parts of the code which control the execution,
  \item \lstinline!exercise04.cc! -- main program,
  \item \lstinline!driver.hh! -- instantiates the necessary PDELab
    classes for solving a linear instationary problem and finally
    solves the problem,
  \item \lstinline!wavefem.hh! -- contains the local operator classes
    \lstinline!WaveFEM! and \lstinline!WaveL2! realizing the spatial
    and temporal residual forms $r(u,v,t)$ and $m(u,v,t)$,
    respectively.
  \end{itemize}

  As in the previous exercises you can control most of the settings
  through the ini-file \lstinline!exercise04.ini!. Get an overview of
  the configurable settings, compile and run \lstinline!exercise04!.

  The program writes output with the extension \lstinline!pvd!. This
  is one of several ways to write VTK output for the instationary
  case, c.f.~the documentation of tutorial03. The \lstinline!pvd!-file
  can be visualized by ParaView and consists of a collection of the
  corresponding \lstinline!vtu!-files. In order to visualize a 1D
  solution, one can apply the ''Plot Over Line '' filter. Note that
  our solution is always given by $u_0$.
\end{Exercise}

\begin{Exercise}{Try various time integrators, in particular the Crank-Nicolson method}
  We want to examine the numerical solution under different time
  discretization schemes, eg. Implicit Euler or Crank-Nicolson. In
  order to change the time discretization scheme you will have to go
  to the file \lstinline!driver.hh! and search for the line
  \begin{lstlisting}
Dune::PDELab::Alexander2Parameter<RF> pmethod;
  \end{lstlisting}

  Recall the previous \lstinline!exercise03! and change the one step
  $\theta$ scheme in a way that corresponds to the Crank-Nicolson
  method.

  \textbf{Note that} you can compare the solutions in ParaView. To do
  that you have to rename the second solution in
  \lstinline!exercise04.ini! and use the "Append Attributes" filter
  before "Plot Over Line". Do not forget to change the parameter
  \lstinline!torder! in \lstinline!exercise04.ini! in order to get the
  correct scheme.

  \textbf{Remark} You can decide which dimension to investigate. A 2D
  simulation will give you nicer pictures at the cost of longer run
  times.
\end{Exercise}


\begin{Exercise}{Explore polynomial degrees greater than $2$ by
    changing the blocking to \lstinline{none}.}
  \textbf{Step 1:} Find the place where your Local Finite Element Maps
  are created and make it possible to use polynomial degree 3

  \begin{lstlisting}
Dune::PDELab::PkLocalFiniteElementMap<GV,DF,double,deg> FEM;
  \end{lstlisting}

  Compile your program and check the results.

  \textbf{Note that} \lstinline!deg! is a static template parameter.

  \textit{The program should throw an assertion error during compilation.}

  \textbf{Step 2:}

  Before we start please have a look at the description of the driver
  in \lstinline!tutorial04! and read the part that describes the
  specification of an ordering when creating product spaces. Using
  fixed block structure in ISTL requires that the number of degrees of
  freedom per entity is constant for each geometry type. This is true
  for polynomial degrees one and two but not for higher polynomial
  degree!

  To avoid segfaults you need to change
  \begin{lstlisting}
Dune::PDELab::istl::VectorBackend<Dune::PDELab::istl::Blocking::fixed> VBE;
  \end{lstlisting}
  to
  \begin{lstlisting}
Dune::PDELab::istl::VectorBackend<Dune::PDELab::istl::Blocking::none> VBE;
  \end{lstlisting}
  in  \lstinline!driver.hh!.
\end{Exercise}

\begin{Exercise}{Changing the Local Operator}
  Now consider the elliptic projection as in Eriksson et
  al. \cite{Eriksson}, namely applying the Laplacian to equation
  \eqref{eq:2b}
  \begin{equation}\label{eq:elip}
    -\Delta \partial_t u_0 + \Delta u_1 = 0,
  \end{equation}
  which has the advantage of energy conservation but requires
  additional smoothness properties.


  The main work is now to change the local operators given in the file
  \lstinline{wavefem.hh}. This is done in several steps:
  \begin{itemize}
  \item Copy the file \lstinline{wavefem.hh} to a new file and rename
    it.
  \item Rename the local operators in the new file, e.g.~change
    \lstinline{WaveL2} to \lstinline{WaveElip}.
  \item Include your new file in \lstinline{exercise04.cc} and change
    the types of \lstinline{LOP} and \lstinline{TLOP} in
    \lstinline{driver.hh}.
  \begin{lstlisting}
// Make instationary grid operator
double speedofsound = ptree.get("problem.speedofsound",1.0);
typedef WaveFEM<FEM> LOP;
LOP lop(speedofsound);
typedef WaveL2<FEM> TLOP;
TLOP tlop;
  \end{lstlisting}
  \end{itemize}

  After these preparations you need to implement the following change:
  $$ \partial_t u_0 - u_1 = 0 \Rightarrow \Delta \partial_t u_0 -
  \Delta u_1 = 0.$$

  \begin{itemize}
  \item In the spatial local operator, change:
    $$ - u_1 = 0 \Rightarrow -  \Delta u_1 = 0$$

    See how it is done for the $\Delta u_0$, we recall the
    corresponding part of \lstinline{WaveFEM::alpha_volume()}:
  \begin{lstlisting}
// integrate both equations
RF factor = ip.weight() * geo.integrationElement(ip.position());
for (size_t i=0; i<lfsu0.size(); i++) {
	r.accumulate(lfsu0,i,c*c*(gradu0*gradphi[i][0])*factor);
	r.accumulate(lfsu1,i,-u1*phihat[i]*factor);
}
  \end{lstlisting}

  \item In the temporal local operator, change:
    $$ \partial_t u_0  = 0 \Rightarrow \Delta \partial_t u_0= 0$$

    See how it is done for the $\Delta u_0$, we recall the
    corresponding part of \lstinline{WaveL2::alpha_volume()}:
  \begin{lstlisting}
// integrate u*phi_i
for (size_t i=0; i<lfsu0.size(); i++) {
	r.accumulate(lfsu0,i,u1*phihat[i]*factor);
	r.accumulate(lfsu1,i,u0*phihat[i]*factor);
}
  \end{lstlisting}

  \item Do not forget to update the \lstinline{jacobian_volume()}
    methods accordingly. As the problem is linear, this should not be
    too difficult.
  \end{itemize}
\end{Exercise}


\begin{Exercise}{Energy conservation}
  If we multiply \eqref{eq:2a} by $u_1$ and \eqref{eq:elip} by $u_0$
  and add, the terms $-(\Delta u_0,u_1 )$ and $(\Delta u_1,u_0 )$
  cancel out, leading to the conclusion that the energy
  $$E(t) = \|u_1\|^2 + \| \nabla u_0\|^2 $$ is constant in time.


  Your task is to check the energy conservation for the elliptic
  projection and Crank-Nicolson in time. You can use the following
  PDELab utilities:
  \begin{lstlisting}
Dune::PDELab::SqrGridFunctionAdapter
Dune::PDELab::integrateGridFunction
Dune::PDELab::DiscreteGridFunctionGradient
Dune::PDELab::SqrGridFunctionAdapter
  \end{lstlisting}

  If you have problems with this task check the online documentation
  \url{https://www.dune-project.org/doxygen/pdelab/master/} or the
  solution.
\end{Exercise}


\textbf{Additional task}\\
If you are done with these exercises, you can play with the initial
conditions. Use the following setting in one space dimension:
\begin{itemize}
\item Change the initial conditions to $\sin(2x)$ (implement it as a
  lambda function)
\item Change the domain size to $[0,\pi]$
\end{itemize}


\bibliographystyle{plain}
\bibliography{exercise04.bib}

\end{document}
\grid
